---
title: "Recriation of '3.2.2 Bad Control' of Angrist and Pischke"
author: "Marcelo"
date: "18/07/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## Simulation 

```{r fig.width=6, fig.height=8, warning=F}
set.seed(12342) # set seed for reproducibility
R <- 1e4      # nr of replications
n <- 1000    # number of observations (set high so to decrease 
              # uncetainty of estimated parameters)

# gen vector of varibles to keep at each replication 
ldiff <- lcausal <- lsbias <- numeric(R)
fitsShort <- fitsLong <- list()
for (r in 1:R) {
    
    y0 <- rnorm(n = n, mean = 8)
    y1 <- rnorm(n = n, mean = 10) # on average, potential oucome is 2 std 
                                  # deviations higher in case of treatment
    δ <- y1 - y0  # effect for each individual
    
    # basically: increases prob of getting blue collar if college attendance
    prob <- 1/10
    
    # potential working outcomes W1 => white collar, W0 -> Blue collar
    W1 <- sample(0:1, size = n, prob = c(prob, 1-prob),replace = T)
    W0 <- sample(0:1, size = n, prob = c(1-prob, prob),replace = T)
    
    ## college is RANDOMLY set
    Ci <- sample(0:1, size = n, replace = T)
    
    # definition of observable variables as defined in pg. 48
    yi <- Ci * y1 + (1 - Ci) * y0
    Wi <- Ci * W1 + (1 - Ci) * W0
    
    # results of apples to oranges from page 49
    diff <- mean(y1[W1 == 1]) - mean(y0[W0 == 1])
    causalColgrad <- mean((y1-y0)[W1==1])
    selecbias <- mean(y0[W1==1]) - mean(y0[W0==1])
    
    ldiff[r] <- diff
    lcausal[r] <- causalColgrad
    lsbias[r] <- selecbias
    
    fitsShort[[r]] <- lm(yi~Ci)
    fitsLong[[r]] <- lm(yi~Ci+Wi)
}

b<-33 # nr of breaks in histogram
{
    par(mfrow=c(3,1),mar=c(3,3,1,1))
    hist(ldiff, b, freq = F); lines(density(ldiff), col="red")
    abline(v=mean(ldiff), col="blue", lwd=2, lty=2)
    hist(lcausal, b, freq = F); lines(density(lcausal), col="red")
    abline(v=mean(lcausal), col="blue", lwd=2, lty=2)
    hist(lsbias, b, freq = F); lines(density(lsbias), col="red")
    abline(v=mean(lsbias), col="blue", lwd=2, lty=2)
}
```

## Data

the variables can be interpreted as:

- yi:  observed outcome
- y1: potential outcome in case of College Attendance
- y0: potential outcome in case of College Non-Attendance
- δ:  Treatmenf effect of each individual
- Ci:  College attendance (0: no, 1:yi es)
- Wi:  Occupation status (0: blue, 1: white collar )
- W1: potential Occupation status in case of College Attendance
- W0: potential Occupation status in case of College Non-Attendance


```{r}
as.data.frame(cbind(
    yi, y1, y0, δ, Ci, Wi, W0, W1)
) %>% round(3) %>% DT::datatable()
```

## Equation 3.2.12

```{r}
mean(yi[Wi==1 & Ci==1]) - mean(yi[Wi==1 & Ci==0])
mean(y1[W1==1 & Ci==1]) - mean(y0[W0==1 & Ci==0])
```

```{r}
mean(y1[W1 == 1 & Ci==1]) - mean(y0[W0 == 1 & Ci==0])
mean(y1[W1 == 1]) - mean(y0[W0==1])
```
not equal due to small sample bias, i assume.


## Regression Comparison


```{r}
# regression of last iteration!
summary(lm(yi ~ Ci))
summary(lm(yi ~ Ci + Wi))
```


## some other stuff of the data


### some observables 

```{r observables}
mean(yi[Ci==1]) # avg observable wages in case of College
mean(yi[Ci==0]) # avg observable wages in case of No College

mean(yi[Wi==1]) # avg observable wages conditional on White Collar
mean(yi[Wi==0]) # avg observable wages conditional on Blue Collar

mean(Ci[Wi==1]) # avg College status conditional on White Collar
mean(Ci[Wi==0]) # avg College status conditional on Blue Collar
```

### some unobservables 

```{r unobsevables}
mean(y0[Ci==1]) # potential wages of college had they not been to college
mean(y1[Ci==0]) # potential wages of no college had they been to college 

mean(y1[Wi==1]) # potential college-wages conditional on occupation status (white collar)
mean(y1[Wi==0]) # potential college-wages conditional on occupation status (blue collar)
mean(y0[Wi==1]) # potential non-college-wages conditional on occupation status (White C.)
mean(y0[Wi==0]) # potential non-college-wages conditional on occupation status (Blue C.)

mean(Ci[W1==1]) # avg observed college attendance conditional on potential occupation status
mean(Ci[W1==0]) # avg observed college attendance conditional on potential occupation status
mean(Ci[W0==1]) # avg observed college attendance conditional on potential occupation status
mean(Ci[W0==0]) # avg observed college attendance conditional on potential occupation status
```

### the four types

```{r}

mean(W1==1 & W0 == 1) # always takers
mean(W1==0 & W0 == 0) # never takers
mean(W1==1 & W0 == 0) # compliers
mean(W1==0 & W0 == 1) # defiers

```


```{r, eval=FALSE, include=FALSE}

#---------------------------------------------------------------
# error 
u1 <- yi - (mean(y0) + (y1-y0) * Ci)
u2 <- y0 - mean(y0)

# effect of College on wages
mean(yi[Ci==1]) - mean(yi[Ci==0])
mean(y1 - y0)


# effect of College on Working status
mean(Wi[Ci==1]) - mean(Wi[Ci==0])
mean(W1 - W0)
```

```{r eq (3.2.12) , eval=FALSE, include=FALSE}
# conditional on Wi == 1: EQ(3.2.12)
mean(yi[Ci==1 & Wi==1]) - mean(yi[Ci==0 & Wi==1])
mean(y1[Ci==1 & W1==1]) - mean(y0[Ci==0 & W0==1])
```

```{r apples to oranges problem, eval=FALSE, include=FALSE}

# by joint independence of y1, w1, y0 w0 on Ci we have:
mean(y1[W1 == 1 & Ci == 1]) - mean(y0[W0 == 1 & Ci == 0])
mean(y1[W1 == 1]) - mean(y0[W0 == 1])

mean((y1-y0)[W1==1]) 

mean(y0[W1 == 1 ]) - mean(y0[W0 == 1]) # selection bias


mean(y0[W1 == 1])
mean(y0[W0 == 1])
mean(y1[W1 == 1])
mean(y1[W0 == 1])

mean(yi[Wi == 1])
mean(yi[Wi == 0])

```

```{r, eval=FALSE, include=FALSE}
mean(W1 == 1 & Ci == 1)
mean(W1 == 1 & Ci == 0)
mean(W1 == 0 & Ci == 1)
mean(W1 == 0 & Ci == 0)

mean((y1-y0)[W1==0]) 
mean((y1-y0)[W1==1]) 

mean((y1-y0)[W0==0]) 
mean((y1-y0)[W0==1]) 

mean((y1-y0)[Ci==1])
mean((y1-y0)[Ci==0])
```

